---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---



###Loading and preprocessing the data
```{r}
activity_data <- read.csv(unz('activity.zip', 'activity.csv'))

# check the data's structure
str(activity_data)

# see some data
tail(activity_data)
```





###What is mean total number of steps taken per day?



For this part of the assignment, you can ignore the missing values in the dataset.


#####1. Make a histogram of the total number of steps taken each day


```{r}
with(na.omit(activity_data), hist(rowsum(steps, date), main='Total Steps In A Day', xlab='Number of Steps'))
```




#####2. Calculate and report the mean and median total number of steps taken per day


```{r}
# calculate in 2 ways for comparison: summary stats, ..
with(na.omit(activity_data), summary(rowsum(steps, date)))

# ...and manual calculation
steps_mean <- with(na.omit(activity_data), mean(rowsum(steps, date)))
steps_mean

steps_median <- with(na.omit(activity_data), median(rowsum(steps, date)))
steps_median
```

**mean** = `r paste(round(steps_mean,2))`

**median** = `r steps_median`




###What is the average daily activity pattern?



#####1. Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)



```{r}
interval_meansteps <- with(na.omit(activity_data), aggregate(steps, list(interval), mean))
names(interval_meansteps) <- c('interval', 'mean_steps')
head(interval_meansteps)


with(interval_meansteps, plot(mean_steps ~ interval, type='l', xlab='Interval', ylab='Average No of Steps'))
```



#####2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?



```{r}
interval_meansteps[which.max(interval_meansteps[,2]),]
```





###Imputing missing values



#####1. Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)
```{r}
nrow(activity_data[is.na(activity_data$steps),])
```



#####2. Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.



Let's fill up with the missing values with **median** in that 5-min interval.





#####3. Create a new dataset that is equal to the original dataset but with the missing data filled in.

```{r}
activity_data_filled <- activity_data
activity_data_filled$steps[is.na(activity_data_filled$steps)] <- with(activity_data_filled, ave(steps, interval, 
   FUN = function(x) median(x, na.rm = TRUE)))[is.na(activity_data_filled$steps)]

# see the result
head(activity_data)
head(activity_data_filled)
```



#####4.Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?




```{r}
with(activity_data_filled, hist(rowsum(steps, date), main='Total Steps In A Day (NA replaced with Median)', xlab='Number of Steps'))
```



As we can see in the histogram above, imputing missing values with the median for the 5-min interval, increases the frequency on the lower end spectrum of number of steps. Consequently, this increase lowers the summary statistics values as confirmed by the table below.



```{r}
with(activity_data_filled, summary(rowsum(steps, date)))
```







###Are there differences in activity patterns between weekdays and weekends?




#####1. Create a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.


```{r}
activity_data_filled$day <- as.factor(ifelse(weekdays(as.Date(activity_data_filled$date)) %in% c('Saturday', 'Sunday'),'weekend','weekday'))
str(activity_data_filled)
```



#####2. Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). See the README file in the GitHub repository to see an example of what this plot should look like using simulated data.


```{r}
# create the data frame
interval_day_mean <- with(activity_data_filled, aggregate(steps, list(interval, day), mean))
names(interval_day_mean) <- c('interval', 'day', 'meansteps')

# see the new data frame
str(interval_day_mean)

# see the data
head(interval_day_mean)

# plot the data
library(lattice)
xyplot(meansteps ~ interval | day, data=interval_day_mean , type='l', layout=c(1,2), xlab='Interval', ylab='Number of Steps')
```


